---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

###Loading and processing the data
```{r functions,  echo=TRUE}
##initialize to use library sqldf
initialize <-function ()
{
  library(sqldf)
  library(ggplot2)
}

##This function return the full path+ file name to be read
getFullPathFileName <-function(pFileName)
{
  return (paste(paste(getwd(),paste("activity", pFileName, sep="/"), sep="/"), ".csv" , sep=""))
}

## Gets the Steps Dt
#This function reads the csv file into a datatable
getDtSteps <- function()
{ 
  if(!exists("dtSteps"))
  {
    dtSteps <<- read.csv(getFullPathFileName('activity'), header=T, sep=",")
  }
  dtSteps$steps <- as.numeric (dtSteps$steps)
  dtSteps$date <- as.Date(dtSteps$date, "%Y-%m-%d")    
  return (dtSteps);  
}



##Gets a datatable without NA values
getDtStepsWithoutNa <- function()
{
  dtSteps <- getDtSteps()  
  return (
sqldf("SELECT 
        *
        FROM dtSteps 
        WHERE steps <>'NA'
        ")
)
}

## Gets the plotting dt
# groupb by day
getPlotDtStepsWithoutNa <-function()
{
  dtSteps<- getDtStepsWithoutNa()
  return( sqldf("SELECT 
        date
        ,sum(steps) as totalSteps
        ,avg(steps) as avgSteps
        ,median(steps) as medianSteps
         
        FROM
        dtSteps
        GROUP BY date
        ")
  )
}

getPlotDtStepsWithoutNaGroupInterval <-function()
{
   dtSteps<- getDtStepsWithoutNa()
   return( sqldf("SELECT 
        interval
        ,sum(steps) as totalSteps
        ,avg(steps) as avgSteps
        ,median(steps) as medianSteps
         
        FROM
        dtSteps
        GROUP BY interval
        ")
   )
}

## @knitr plotHistogramWithoutNa
plotHistogramWithoutNa <-function()
{
  dtPlot <- getPlotDtStepsWithoutNa()
  #dtPlot <- sqldf("select date, totalSteps from dtPlot")
  #qplot(dtPlot$date, dtPlot$sumSteps) +geom_histogram()
  ggplot(data=dtPlot, aes(x=date, y=totalSteps)) + 
     geom_bar(stat="identity", colour="black") +
     xlab("date") +
     ylab("Total Steps") +
     ggtitle("Total steps per day")  
 
}

#This is used to fill missing values
getDtMedianStepsPerInterval <-function()
{
   dtSteps <- getDtSteps()
   return (sqldf("select interval, median(steps) as steps from dtSteps WHERE steps <>'NA'  group by interval"))
}

#insert missing values
getDtFilledMissingValues <- function()
{
   dtMedianMissingSteps <- getDtMedianStepsPerInterval()
   dtMissingSteps<- sqldf("select step1.*, 
                           step2.steps as missingStep 
                           FROM dtSteps step1 
                           left join dtMedianMissingSteps step2 on step1.interval = step2.interval")
   
   return(sqldf("select 
                dtMissingSteps.*
                , case  
                  when steps is null then missingStep 
                  else steps end as adjStep 
                from dtMissingSteps"))
}

initialize()
dtSteps<- getDtSteps()
dtStepsWithoutNa<- getDtStepsWithoutNa()
dtPlotStepsWithoutNa <- getPlotDtStepsWithoutNa()

```

## What is mean total number of steps taken per day?

```{r avgSteps}
 ggplot(data=dtPlotStepsWithoutNa, aes(x=date, y=totalSteps)) + 
     geom_bar(stat="identity", colour="black") +
     xlab("date") +
     ylab("Total Steps") +
     ggtitle("Total steps per day")  

```

### Mean steps per day:
```{r meanSteps}
mean(dtPlotStepsWithoutNa$totalSteps)
```

###Median stepsper day
```{r medianSteps}
median(dtPlotStepsWithoutNa$totalSteps)  
```

## What is the average daily activity pattern?
```{r activityPattern}
  dtStepsWithoutNa <- getPlotDtStepsWithoutNaGroupInterval()
  dtStepsWithoutNa$avgSteps <-mean(dtPlotStepsWithoutNa$totalSteps)
 

ggplot(data=dtStepsWithoutNa, aes(x=interval, y=totalSteps)) + 
     geom_line(colour="blue", linetype="solid") +
     xlab("interval") +
     ylab("Steps") +
     ggtitle("Total steps per interval") 
     
```

Interval with the highest number of steps:
```{r maxNumberSteps}
#selects the interval corresponding to the max number of steps
 sqldf("select  interval  from dtStepsWithoutNa WHERE totalSteps = (select max(totalSteps) from dtStepsWithoutNa)")
```


## Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r totalNumberMissingValues}
#get dataste without NA
dtStepsWithoutNa <- getDtStepsWithoutNa()
#subtracts the number of rows in the dataset withoutNA from full dataset
nrow(dtSteps) - nrow(dtStepsWithoutNa)
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

--> Used median steps per interval. This is done in function getDtMedianStepsPerInterval()

Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r fillingMissingValues}
## getDtFilledMissingValues() adds a new column (adjStep) that
## contains steps (if <> NA) or the median(steps) for that interval ##oterhwise
dtMissingSteps <-getDtFilledMissingValues()
```

## Are there differences in activity patterns between weekdays and weekends?
